Modal Abstractions for Virtualizing Memory Addresses

### 2024-11-12 programlogic.tex
compression, clarifying relation to syeda et al

\subsection{Address-Space Management}
\label{sec:aspacemanagement}
So far, we have introduced logical abstractions for a single address space, but VMMs
 handle more than one address-space,
and doing so requires a way to talk about other address spaces, and means to switch address spaces.
\begin{figure}
\footnotesize
\begin{mathpar}
    \mbox{$[r](\mathsf{P}) \;: \mathsf{vProp} \; \Sigma \; \stackrel{\triangle}{=} \; \lambda \_,\; \mathsf{P}~\mathsf{r} \vspace{0.5em}$}
    \and
    \mbox{$\textsf{Fact P} \stackrel{\triangle}{=} \;\forall \; ,r \; r' \ldotp  \textsf{P r} \dashv\vdash \textsf{P r'}$}
    \and
    \inferrule*{ }{\textsf{Fact}\;[r](\textsf{P}) }
    \and
    \inferrule*{ }{\textsf{Fact}\;(\textsf{r}\;\mapsto_{\textsf{r}}\{q\}\;\textsf{rv})}
    \and
    \inferrule{ }{\textsf{Fact}\;(\textsf{w} \mapsto_{\textsf{p}} \{q\} \textsf{ v} )}
    \and
    \inferrule{ }{(\textsf{P} \vdash \textsf{Q}) \vdash  ([r](\textsf{P}) \vdash  [r](\textsf{Q}))}
    \and
    \inferrule{ }{[r](\textsf{P} \ast \textsf{Q}) \dashv\vdash ([r](\textsf{P}) \ast [r](\textsf{Q})}\vspace{0.5em}
    \and
    \inferrule{ }{[r](\textsf{P}\land \textsf{Q}) \dashv\vdash [r](\textsf{P})\land [r](\textsf{Q})}
    \and
    \inferrule{ }{[r](\textsf{P}\lor \textsf{Q}) \dashv\vdash [r](\textsf{P})\lor [r](\textsf{Q})}
    \and
    \inferrule{ }{\textsf{Fact P} \vdash  [r](\textsf{P}) \dashv\vdash (\textsf{P})} %(\textsf{vProp }\Sigma))\\
\end{mathpar}
\vspace{-1em}
  \caption{Other-space Modality and Its Laws}
  \label{fig:modaldef}
\vspace{-1em}
  \end{figure}
Figure \ref{fig:modaldef} gives the definition of our modal operator for asserting truth of a modal
(address-space-contingent) assertion \emph{in another address space}, which we call
the \emph{other-space} modality. The definition itself is not
particularly surprising --- as our modal assertions are semantically predicates on a page table root (physical)
address, the assertion $[r](P)$ is a modal assertion which ignores the (implicit) current page table root,
and evaluates the truth of $P$ as if $r$ were the page table root. 
The novelty here is not in the details of the definition, but in recognizing that this is the right way to deal with
multiple address spaces, and working out how to support interaction of multiple address spaces (discussed in the next section).
% This simple definition is 
% what we need in order for assertions in one address space to talk about assertions in another.
% The inspiration for this assertion comes from hybrid logic~\cite{blackburn1995hybrid,areces2001hybrid,goranko1996hierarchies,gargov1993modal},
% which as mentioned earlier uses as modalities the names of worlds in a Kripke model to state
% that an assertion is true in that specific world.
% Conceptually the address space rooted at some physical address (intended as the start of an L4 table)
% is a distinct world, and it makes sense to name it by referring to the root of the
% supporting page table structure. However, rather than explicitly building a Kripke model for our
% assertions, we inherit them by lifting the model of our ambient logic, Iris.
We can prove that this modality follows certain basic laws: that its truth is independent of the address
space in which it is considered,  it distributes over various logical connectives, and it follows the rule of
consequence.
We call \textsf{vProp} assertions whose truth is independent of the current address space
\textsf{Fact}s; these include other-space assertions, physical memory points-tos, and register assertions.
\textsf{Fact}s can move in and out of other-space modalities freely.
\looseness=-1

@@

\subsection{Subtleties of Changing Address Spaces Using Modalities}
\todo[inline]{Colin will push this to an appendix, and leave just a short comment of a few lines here with a forward ref}
% \subsection{Considering Hoare Doubles in the Context of Switching Address-Space}
% \subsubsection{Issues with Using Hoare Triples in Address-Space Specification}
\label{sec:issues}
Operating systems on x86-64 change address spaces by updating \lstinline|cr3|.
Conceptually, we expect a proof rule for such an update to take move resources in and out of modalities for
the old and new address spaces. This suggests a rule like the following, which captures the intuition of the
address space change, but is subtly unsound under common assumptions.
% An important subtlety arises with supporting \lstinline|mov|s into \lstinline|cr3|. Consider the hypothetical rule
\begin{mathpar}
\inferrule*[left=Broken]{ }{
  \{P \ast \textsf{cr3}\mapsto_{r} \;r_1 \ast r \mapsto_{r} \;r_2 \ast [r_2](Q)\}
  \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
  \{[r_1](P) \ast \textsf{cr3} \mapsto_{r} r_2 \ast r \mapsto_{r} r_2 \ast Q\}
}
\end{mathpar}
This rule takes all assertions not under a modality ($P$), and moves them under a modality for the old address space
(which is where they were valid), while pulling assertions for the new, target address space out of a corresponding modality.
(For clarity, we also omit the address space invariants, which must also be present.)
% which is slightly simplified (by dropping per-address-space resources) for clarity in making the following point.
% This rule captures the intuitive change of address space in a Hoare triple.
Unfortunately, as stated, this interacts quite poorly with the traditional frame
rule.
If we abbreviate the pre- and post-conditions above as \textsf{Pre} and \textsf{Post}:
\begin{mathpar}
  \inferrule*[right=Frame]{
    \inferrule*[right=Broken]{ }{
    \{\textsf{Pre}\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{\textsf{Post}\}
    }
  }{
    \{a\mapsto_\mathsf{v} x \ast \textsf{Pre}\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{a\mapsto_\mathsf{v} x \ast \textsf{Post}\}
  }
\end{mathpar}
Notice that both the precondition and postcondition assert that $a\mapsto_\mathsf{v} x$ in the current address space, but we have no basis for concluding that address translation is preserved by the change of address space. So this derivation clearly leads to an unsound conclusion. 
The essential problem is that the frame rule is motivated by local reasoning about local updates, but
a switch of address space is a \emph{global} change that may invalidate information about virtual addresses.
Thus framing around \lstinline|cr3| updates is unsound, which has important impacts on our formal development.

Switching to Hoare doubles resolves this problem because an under-appreciated subtlety of Hoare doubles 
is that typically \emph{there is no primitive frame rule}. 
Instead each verification essentially includes a local frame that it passes to the next instructions
(think continuation-passing style), giving each overall rule a \emph{global} (rather than local) precondition.
For most rules this is not that important, but it does permit rules that have global effects on their preconditions.

% This is then how we justify our actual rule for \lstinline|cr3| updates, which is \emph{almost}
This leads to our actual rule for \lstinline|cr3| updates, which is \emph{almost}
\begin{mathpar}
\inferrule*[left=ChangeAddressSpace]{
  \{[r_1](P) \ast cr3 \mapsto_{r} r_2 \ast r \mapsto_{r} r_2 \ast Q\}\;\overline{is}
}{
  \{P \ast cr3 \mapsto_{r} r_1 \ast r \mapsto_{r} r_2 \ast [r_2](Q)\}\;
  \texttt{mov}~\texttt{\%cr3},~r;\;\overline{is}
  %\lstinline|mov %cr3, r| 
}
\end{mathpar}
(Again, the actual rule in Section \ref{sec:selected_rules} includes the address space invariants.)
Because the precondition on this rule is global, we avoid issues with framing. 
A more typical frame rule can be recovered from Hoare doubles in the typical way~\cite{Chlipala2013Bedrock} 
with the added restriction that \lstinline|cr3| is restored to its pre-framing value before un-framing (in this case, the framed-out assertion is effectively
stashed under an other-space assertion for the initial address space, and pulled out as part of restoring).

% If we wanted to consider a frame rule that would work for this logic, we could consider:
% \begin{mathpar}
%   \inferrule*[left=Cr3Frame]{
%     \{P\ast cr3=v\}\;C\;\{ Q \ast cr3=v\}
%   }{
%     \{R\ast P\ast cr3=v\}\;C\;\{ R \ast Q \ast cr3=v\}
%   }
% \end{mathpar}
% By demanding that \lstinline|cr3| be held constant (or rather, at least restored to its original value) we could frame almost traditionally. In particular, this rule would work with framing around calls that might lead to address space switches, such as calling blocking operations in the kernel.
% Attempting to derive a frame rule as some work using Hoare doubles does~\cite{Chlipala2013Bedrock}
% would enforce this restriction.

% Another useful perspective on this might be the dynamic frames~\cite{parkinson2011relationship} 
% view: virtual points-to assertions are self-stable \emph{except} for changes in \lstinline|cr3|, so framing would then naturally require other means of holding \lstinline|cr3| constant (or saving and restoring it).
% Virtual points-to assertions could be made self-stable by also giving them partial ownership over \lstinline|cr3| assertions, but this would require explicitly plumbing that ownership from \emph{all} assertions back to any place an address space change might occur; this would seem to be a far graver loss of modularity than this extra quirk in framing discussions.

\subsection{Selected Logical Rules}
\label{sec:selected_rules}
Per the discussion in Section \ref{sec:issues}, we define our logic using Hoare doubles:\footnote{This
omits some low-level Iris details (stuckness, observations) that play no meaningful
role in our development.}
  \[
  \begin{array}{l}
    %\textsf{wpd\_def e s E1 } \Phi \;\mathsf{ rtv } : \textsf{iProp }\Sigma := \\
    \{ \Phi \}_\mathsf{ rtv }\;\textsf{e} : \textsf{iProp }\Sigma := 
   % \qquad
   ((\textsf{cr3} \mapsto_{\textsf{r}} \textsf{rtv} \ast \Phi) \textsf{ rtv}) \wand \textsf{WP e } \{\_, \textsf{True} \}
    \end{array}
  \]
Our Hoare doubles $\{\Phi\}_\textsf{rtv}\;\textsf{e}$ state that the expression (i.e., sequence of instructions)
\textsf{e} are safe to execute (will not fault)
when executed with \textsf{vProp} precondition $\Phi\ast\textsf{cr3}\mapsto_{\textsf{r}} \textsf{rtv}$.
\textsf{WP} is Iris's own weakest precondition modality, unmodified~\cite{jung2018iris}.
Making \textsf{rtv} an explicit parameter to the double, rather than simply using register assertions
directly solves a technical problem with ensuring that the page table root used to evaluate
the \textsf{vProp} (i.e., evaluating the assertion in the \emph{current}) address space
is feasible.\footnote{Consider the difficulty of selecting the correct page table root value from an arbitrary
opaque $\Phi$, which may even existentially quantify the page table root. An alternative is to
require $\Phi$ to have a syntactic form where we can directly extract the value of \lstinline|cr3|,
but this makes using Iris Proof Mode (IPM)~\cite{Krebbers:2017:IPH:3009837.3009855} with \textsf{vProps}
  difficult; IPM works for any type matching the signature of an Iris \lstinline|bi|, which includes
  \textsf{vProp}s, but manually guiding IPM to put an assertion in a specific position over and over adds
  significant proof burden.
}

The rest of this section describes specifications of some selected \textsf{AMD64} instructions 
in our logic. 
These rules and others (e.g., including accessing memory at an instruction-specified offset from a register
value, which is common in most ISAs)
can be found in our artifact.
Each rule in Figure \ref{fig:wpdamd}, 
is annotated with a root (i.e., \lstinline|cr3|) address value (\textsf{rtv}), 
under which the resources mentioned in the specification are valid.
In general, we use metavariables $\textsf{r}_s$ and $\textsf{r}_d$ to specify source and destination registers
for each instruction, and prefix various register value variables with \textsf{rv}.
We sometimes use $\textsf{r}_a$ to emphasize when a register is expected to hold an address used
for memory access, though the figure also uses typical assembler conventions of specifying
memory access operands by bracketing the register holding the memory address.
Standard for Hoare doubles, there is a frame resource $P$ in each rule for passing resources
not used by the first instruction in sequence through to subsequent instructions.
Our rules include tracking of each instruction's memory address to track \lstinline|rip| updates, which is critical
for control transfer instructions. Our development also includes handling of the \lstinline|rflags| register updates from arithmetic instructions.
Most rules are otherwise standard (e.g., \lstinline|mov| between register, etc.), with Figure \ref{fig:wpdamd} showing the rules
most unique to our development.
\looseness=-1

@@

@@

\paragraph{Updating \lstinline|cr3|} 
Unlike other rules, \TirNameStyle{WriteToRegCtlFromRegModal} updates the root address of the 
address-space determining the validity of resources, from $\rtv$ before the
\lstinline|mov| to $\textsf{rvs}$ afterwards. The global effects of this rule are reflected in moving
bare assertions under an other-space modality for $\rtv$, and moving the new address space's assertions out of
the corresponding modality.
% From that rule, with one step outside the \textsf{vProp} logic (so it is not quite a derived rule within
% our embedded logic), we can derive 
% \TirNameStyle{WriteToRegCtlFromRegModal},
% which is the (full) Hoare double version of the address space switch rule proposed in Section \ref{sec:issues}.

>>>>

\subsection{Address-Space Management}
\label{sec:aspacemanagement}
% So far, we have introduced logical abstractions for a single address space, but VMMs
Real VMMs must
 handle more than one address-space.
Doing so requires a way to talk about other address spaces, and means to switch address spaces.
\begin{figure}
\footnotesize
\begin{mathpar}
    \mbox{$[r](\mathsf{P}) \;: \mathsf{vProp} \; \Sigma \; \stackrel{\triangle}{=} \; \lambda \_,\; \mathsf{P}~\mathsf{r} \vspace{0.5em}$}
    \and
    \mbox{$\textsf{Fact P} \stackrel{\triangle}{=} \;\forall \; ,r \; r' \ldotp  \textsf{P r} \dashv\vdash \textsf{P r'}$}
    \and
    \inferrule*{ }{\textsf{Fact}\;[r](\textsf{P}) }
    \and
    \inferrule*{ }{\textsf{Fact}\;(\textsf{r}\;\mapsto_{\textsf{r}}\{q\}\;\textsf{rv})}
    \and
    \inferrule{ }{\textsf{Fact}\;(\textsf{w} \mapsto_{\textsf{p}} \{q\} \textsf{ v} )}
    \and
    \inferrule{ }{(\textsf{P} \vdash \textsf{Q}) \vdash  ([r](\textsf{P}) \vdash  [r](\textsf{Q}))}
    \and
    \inferrule{ }{[r](\textsf{P} \ast \textsf{Q}) \dashv\vdash ([r](\textsf{P}) \ast [r](\textsf{Q})}\vspace{0.5em}
    \and
    \inferrule{ }{[r](\textsf{P}\land \textsf{Q}) \dashv\vdash [r](\textsf{P})\land [r](\textsf{Q})}
    \and
    \inferrule{ }{[r](\textsf{P}\lor \textsf{Q}) \dashv\vdash [r](\textsf{P})\lor [r](\textsf{Q})}
    \and
    \inferrule{ }{\textsf{Fact P} \vdash  [r](\textsf{P}) \dashv\vdash (\textsf{P})} %(\textsf{vProp }\Sigma))\\
\end{mathpar}
\vspace{-1em}
  \caption{Other-space Modality and Its Laws}
  \label{fig:modaldef}
\vspace{-1em}
  \end{figure}
Figure \ref{fig:modaldef} gives the definition of our modal operator for asserting truth of a modal
(address-space-contingent) assertion \emph{in another address space}, which we call
the \emph{other-space} modality. The definition itself is not
particularly surprising --- as our modal assertions are semantically predicates on a page table root (physical)
address, the assertion $[r](P)$ is a modal assertion which ignores the (implicit) current page table root,
and evaluates the truth of $P$ as if $r$ were the page table root. 
The novelty here is not in the details of the definition, but in recognizing that this is the right way to deal with
multiple address spaces, and working out how to support interaction of multiple address spaces (discussed in the next section).
% This simple definition is 
% what we need in order for assertions in one address space to talk about assertions in another.
% The inspiration for this assertion comes from hybrid logic~\cite{blackburn1995hybrid,areces2001hybrid,goranko1996hierarchies,gargov1993modal},
% which as mentioned earlier uses as modalities the names of worlds in a Kripke model to state
% that an assertion is true in that specific world.
% Conceptually the address space rooted at some physical address (intended as the start of an L4 table)
% is a distinct world, and it makes sense to name it by referring to the root of the
% supporting page table structure. However, rather than explicitly building a Kripke model for our
% assertions, we inherit them by lifting the model of our ambient logic, Iris.
We can prove that this modality follows certain basic laws: that its truth is independent of the address
space in which it is considered,  it distributes over various logical connectives, and it follows the rule of
consequence.
We call \textsf{vProp} assertions whose truth is independent of the current address space
\textsf{Fact}s; these include other-space assertions, physical memory points-tos, and register assertions.
\textsf{Fact}s can move in and out of other-space modalities freely.
\looseness=-1

@@

\subsection{Selected Logical Rules}
\label{sec:selected_rules}
% Per the discussion in Section \ref{sec:issues}, w
As common for assembly-level verification, we define our logic using Hoare doubles:\footnote{This
omits some low-level Iris details (stuckness, observations) that play no meaningful
role in our development.}
  \[
  \begin{array}{l}
    %\textsf{wpd\_def e s E1 } \Phi \;\mathsf{ rtv } : \textsf{iProp }\Sigma := \\
    \{ \Phi \}_\mathsf{ rtv }\;\textsf{e} : \textsf{iProp }\Sigma := 
   % \qquad
   ((\textsf{cr3} \mapsto_{\textsf{r}} \textsf{rtv} \ast \Phi) \textsf{ rtv}) \wand \textsf{WP e } \{\_, \textsf{True} \}
    \end{array}
  \]
Our Hoare doubles $\{\Phi\}_\textsf{rtv}\;\textsf{e}$ state that the expression (i.e., sequence of instructions)
\textsf{e} are safe to execute (will not fault)
when executed with \textsf{vProp} precondition $\Phi\ast\textsf{cr3}\mapsto_{\textsf{r}} \textsf{rtv}$.
\textsf{WP} is Iris's own weakest precondition modality, unmodified~\cite{jung2018iris}.
Making \textsf{rtv} an explicit parameter to the double, rather than simply using register assertions
directly 
makes it possible to ensure ownership of the \lstinline|cr3| register and its value is accounted for
while avoiding some technical headaches with trying to enforce that $\Phi$ itself contains that.
% solves a technical problem with ensuring that the page table root used to evaluate
% the \textsf{vProp} (i.e., evaluating the assertion in the \emph{current}) address space
% is feasible.
% \footnote{Consider the difficulty of selecting the correct page table root value from an arbitrary
% opaque $\Phi$, which may even existentially quantify the page table root. An alternative is to
% require $\Phi$ to have a syntactic form where we can directly extract the value of \lstinline|cr3|,
% but this makes using Iris Proof Mode (IPM)~\cite{Krebbers:2017:IPH:3009837.3009855} with \textsf{vProps}
%   difficult; IPM works for any type matching the signature of an Iris \lstinline|bi|, which includes
%   \textsf{vProp}s, but manually guiding IPM to put an assertion in a specific position over and over adds
%   significant proof burden.
% }

The rest of this section describes specifications of three key \textsf{AMD64} instructions 
in our logic. 
These rules and others (e.g., including accessing memory at an instruction-specified offset from a register
value, which is common in most ISAs)
can be found in our artifact.
Each rule in Figure \ref{fig:wpdamd}, 
is annotated with a root (i.e., \lstinline|cr3|) address value (\textsf{rtv}), 
under which the resources mentioned in the specification are valid.
In general, we use metavariables $\textsf{r}_s$ and $\textsf{r}_d$ to specify source and destination registers
for each instruction, and prefix various register value variables with \textsf{rv}.
We sometimes use $\textsf{r}_a$ to emphasize when a register is expected to hold an address used
for memory access, though the figure also uses typical assembler conventions of specifying
memory access operands by bracketing the register holding the memory address.
Standard for Hoare doubles, there is a frame resource $P$ in each rule for passing resources
not used by the first instruction in sequence through to subsequent instructions.
Our rules include tracking of each instruction's memory address to track \lstinline|rip| updates, which is critical
for control transfer instructions. Our development also includes handling of the \lstinline|rflags| register updates from arithmetic instructions.
Most rules are otherwise standard (e.g., \lstinline|mov| between register, etc.), with Figure \ref{fig:wpdamd} showing the rules
most unique to our development.
\looseness=-1

@@

\vspace{-1em}

@@

\paragraph{Updating \lstinline|cr3|} 
Unlike other rules, \TirNameStyle{WriteToRegCtlFromRegModal} updates the root address of the 
address-space determining the validity of resources, from $\rtv$ before the
\lstinline|mov| to $\textsf{rvs}$ afterwards. The global effects of this rule are reflected in moving
bare assertions under an other-space modality for $\rtv$, and moving the new address space's assertions out of
the corresponding modality.
Note that the \emph{global} aspect is important, as a na\"ive frame rule would be unsound for updates to \lstinline|cr3|,
because one could frame out assertions in one address space, switch address spaces, and bring those assertions from the \emph{old}
address space back into the \emph{new} address space, where they may not hold. Appendix \ref{sec:issues} gives more details.
% From that rule, with one step outside the \textsf{vProp} logic (so it is not quite a derived rule within
% our embedded logic), we can derive 
% \TirNameStyle{WriteToRegCtlFromRegModal},
% which is the (full) Hoare double version of the address space switch rule proposed in Section \ref{sec:issues}.

### 2024-11-12 relwork.tex
compression, clarifying relation to syeda et al

They also define their virtual points-to assertions such that a virtual points-to $p\mapsto_\mathsf{v} a$ owns the full 
lookup path to virtual address $p$. This means that given two virtual points-to assertions at the same time, such as 
$p\mapsto_\mathsf{v}a \ast p'\mapsto_\mathsf{v}b$, the memory locations traversed to translate $p$ and $p'$ must be disjoint. 
This means the logic has a peculiar limit on how many virtual points-to assertions can coexist in a proof. Since page tables 
fan out, the bottleneck is the number of entries in the root table. For their 32-bit ARMv6 example, the top-level address is 
still 4Kb (4096 bytes), and each entry (consumed entirely by a virtual points-to in their scheme) is 4 bytes, so they have a 
maximum of 1024 virtual points-tos in their ARMv6 configuration. Any assertion which implies more than that number
of virtual addresses are mapped implies false in their logic.
(They do formulate their logic over an abstract model, but every architecture would incur a similar limitation;
Na\"ively transferring their model to x86-64 4-level tables would yield a limit of 512 assertions (also a 4Kb root page, 
but 8-byte entries).
Our definitions make use of fractional permissions throughout; Figure \ref{fig:strongvirtualpointsto}'s definition
of \lstinline|L4_L1_PointsTo| ellides the specific fractions used, but it in fact asserts 1/512 ownership of
the L1 entry, 1/($512^2$) of the L2 entry, and so on, so each entry may map the appropriate number of machine words.

As noted earlier, Kolanski and Klein's logics, by collocating both the physical ownership of the page table walk
as part of the virtual points-to itself, preempt support for changes to page tables which do not actually affect 
address translation. We address this by moving from Figure \ref{fig:strongvirtualpointsto}'s definition (essentially Kolanski and Klein's
with fractional sharing of intermediate table entries) to Figure \ref{fig:virtualpointstosharing}'s.

Kolanski and Klein do verify code to map in a new page by installing an L1 page table entry, akin to our Figure \ref{fig:mapping_code}.
However, our logic treatment of mapping goes beyond theirs. 
Kolanski and Klein must unfold machine semantics at one point in their proof of mapping correctness~\cite[p.~27]{kolanski08vstte}, while
our logic permits us to conduct the proof entirely in separation logic.
Their proofs also do not address
converting physical addresses of page table entries to virtual addresses that can be used to access them.
Their original low-level (pseudo-assembly) proof assumes the correct virtual address already exists~\cite{kolanski08vstte},
and their subsequent C-level proof~\cite{kolanski09tphols} axiomatizes a function akin to our \lstinline|ensure_L1|.
While full proof of \lstinline|ensure_L1|'s page table walk is ongoing work, we have verified that virtual-to-physical
translation can be done in our logic with appropriate kernel invariants (Section \ref{sec:traversing}), while
Kolanski and Klein's model may not be able to do so --- because the virtual points-to owns the physical resources associated with
intermediate page table entries, it appears impossible to have a virtual points-to for a page table entry that is part of the
page table walk for another virtual address.

The other major distinction is that Kolanski and Klein have no explicit accounting for other address spaces.
Their logic does not deal with change of address space, and has no way to assert that certain facts hold
in another address space.
They verify only one address space manipulation: mapping a single unmapped page into the current address space (in both papers).
We verify this, as well as a change-of-address-space, which requires us to introduce assertions for talking
about other address spaces (we must know, for example, that the precondition of the code after the change must be true
in the \emph{other} address space), and to deal with the fact that the standard frame rule
for separation logic is unsound in the presence of address space changes and address-space-contingent assertions.
Our approach in this paper uses modalities to distinguish virtual-address-based assertions that hold only in specific 
address spaces, making it possible to manipulate other address spaces, and equally critically, to \emph{change} address 
spaces while reasoning about correctness.

@@

As noted earlier, our other-space modality comes from ideas in hybrid logic~\cite{areces2001hybrid,blackburn1995hybrid,gargov1993modal,goranko1996hierarchies},
where modalities are indexed by \emph{nominals} which are names for specific individual states in a Kripke model.
We are aware of only two prior works combining these ideas with program logics, one showing that traditional
nominals extends the expressive power of separation logic~\cite{brotherston2014parametric}, another
using nominals to refer to states of other actors in an actor language~\cite{gordon2019modal}.

% As noted throughout the paper, the inspiration for our other-space modality comes from hybrid logic~\cite{areces2001hybrid,blackburn1995hybrid,gargov1993modal,goranko1996hierarchies},
% where modalities are indexed by \emph{nominals} which are names for specific individual states in a Kripke model.
% We are aware of only two prior works combining hybrid logics with program logics specifically. 
% Brotherston and Villard~\cite{brotherston2014parametric} demonstrated that may properties true of various 
% separation logics are not definable in boolean \BI (\BBI), and showed that a hybrid extension \HyBBI allows
% most such properties to be defined (e.g., the fact that separating conjunction is cancellative is unprovable 
% in boolean \BI, but provable in \HyBBI). There, nominals named resources 
% (roughly, but not exactly, heap fragments). 
% Gordon~\cite{gordon2019modal} described a use of hybrid logic in the verification of actor programs, 
% where nominals named the local state of individual actors (with such assertions stabilized with a 
% rely/guarantee approach). 
Beyond these, there is limited work on the interaction of hybrid logic with substructural logics, in restricted forms
that do not affect expressivity~\cite{despeyroux2014hybrid,chaudhuri2019hybrid}.
% Primarily there is a line of work on hybrid linear logic (\HyLL)~\cite{despeyroux2014hybrid}, 
% originally used as a way to more conveniently express aspects of transition systems in linear logic. 
% However, \HyLL's proof rules offer no non-trivial interactions with multiplicative connectives 
% (every \HyLL proof can in fact be embedded into regular linear logic~\cite{chaudhuri2019hybrid}, 
% unlike Brotherston and Villard's \HyBBI, which demonstrably increases expressive power over its base \BBI.

>>>>

They also define their virtual points-to assertions such that a virtual points-to $p\mapsto_\mathsf{v} a$ owns the full 
lookup path to virtual address $p$. This means that given two virtual points-to assertions at the same time, such as 
$p\mapsto_\mathsf{v}a \ast p'\mapsto_\mathsf{v}b$, the memory locations traversed to translate $p$ and $p'$ must be disjoint. 
This means the logic has a peculiar limit on how many virtual points-to assertions can coexist in a proof. Since page tables 
fan out, the bottleneck is the number of entries in the root table. For their 32-bit ARMv6 example, the top-level address is 
still 4Kb (4096 bytes), and each entry (consumed entirely by a virtual points-to in their scheme) is 4 bytes, so they have a 
maximum of 1024 virtual points-tos in their ARMv6 configuration. Any assertion which implies more than that number
of virtual addresses are mapped implies false in their logic.
(They do formulate their logic over an abstract model, but every architecture would incur a similar limitation.)
% Na\"ively transferring their model to x86-64 4-level tables would yield a limit of 512 assertions (also a 4Kb root page, 
% but 8-byte entries).
Our definitions make use of fractional permissions throughout; Figure \ref{fig:strongvirtualpointsto}'s definition
of \lstinline|L4_L1_PointsTo| ellides the specific fractions used, but it in fact asserts 1/512 ownership of
the L1 entry, 1/($512^2$) of the L2 entry, and so on, so each entry may map the appropriate number of machine words.
\looseness=-1

As noted earlier, by collocating both the physical ownership of the page table walk
as part of the virtual points-to itself these logics preempt support for changes to page tables which do not actually affect 
address translation. We address this by moving from Figure \ref{fig:strongvirtualpointsto}'s definition (essentially Kolanski and Klein's
with fractional sharing of intermediate table entries) to Figure \ref{fig:virtualpointstosharing}'s.
\looseness=-1

Kolanski and Klein do verify code to map in a new page by installing an L1 page table entry, akin to our Figure \ref{fig:mapping_code}.
However, our logic treatment of mapping goes beyond theirs. 
Kolanski and Klein must unfold machine semantics at one point in their proof of mapping correctness~\cite[p.~27]{kolanski08vstte}, while
our logic permits us to conduct the proof entirely in separation logic.
Their proofs also do not address
converting physical addresses of page table entries to virtual addresses that can be used to access them.
Their original low-level (pseudo-assembly) proof assumes the correct virtual address already exists~\cite{kolanski08vstte},
and their subsequent C-level proof~\cite{kolanski09tphols} axiomatizes a function akin to our \lstinline|ensure_L1|.
While full proof of \lstinline|ensure_L1|'s page table walk is ongoing work, we have verified that virtual-to-physical
translation can be done in our logic with appropriate kernel invariants (Section \ref{sec:traversing}), while
Kolanski and Klein's model may not be able to do so --- because the virtual points-to owns the physical resources associated with
intermediate page table entries, it appears impossible to have a virtual points-to for a page table entry that is part of the
page table walk for another virtual address.
\looseness=-1

The other major distinction is that Kolanski and Klein have no explicit accounting for other address spaces.
Their logic does not deal with change of address space, and has no way to assert that certain facts hold
in another address space.
They verify only one address space manipulation: mapping a single unmapped page into the current address space (in both papers).
We verify this, as well as a change-of-address-space, which requires us to introduce assertions for talking
about other address spaces (we must know, for example, that the precondition of the code after the change must be true
in the \emph{other} address space), and to deal with the fact that the standard frame rule
for separation logic is unsound in the presence of address space changes and address-space-contingent assertions.
Our approach in this paper uses modalities to distinguish virtual-address-based assertions that hold only in specific 
address spaces, making it possible to manipulate other address spaces, and equally critically, to \emph{change} address 
spaces while reasoning about correctness.
The only other attempt to verify a change of address space was by \citet{syeda2018program,syeda2020formal}, whose focus was on TLB maintenance (which we
do not model). However, they verified only the 4 instructions to switch address spaces and update the TLB on an ARM processor, in isolation.
The specification they proved for those instructions did not address program invariants that may be valid in one address space and not the other,
so is not flexible enough to extend directly to a full context switching primitive as in Figure \ref{fig:swtch}.
Of course as we do not address TLB management (see Appendix \ref{apdx:tlb}), the right eventual solution would be to combine our work
and theirs, which we leave to future work.

@@

As noted earlier, our other-space modality derives from hybrid logic~\cite{areces2001hybrid,blackburn1995hybrid,gargov1993modal,goranko1996hierarchies},
where modalities are indexed by \emph{nominals} which are names for specific individual states in a Kripke model.
Little prior work combines these ideas with program logics: \citet{brotherston2014parametric} show that traditional
nominals extends the expressive power of separation logic, and \citet{gordon2019modal}
uses nominals to refer to states of other actors in an actor language.
% As noted throughout the paper, the inspiration for our other-space modality comes from hybrid logic~\cite{areces2001hybrid,blackburn1995hybrid,gargov1993modal,goranko1996hierarchies},
% where modalities are indexed by \emph{nominals} which are names for specific individual states in a Kripke model.
% We are aware of only two prior works combining hybrid logics with program logics specifically. 
% Brotherston and Villard~\cite{brotherston2014parametric} demonstrated that may properties true of various 
% separation logics are not definable in boolean \BI (\BBI), and showed that a hybrid extension \HyBBI allows
% most such properties to be defined (e.g., the fact that separating conjunction is cancellative is unprovable 
% in boolean \BI, but provable in \HyBBI). There, nominals named resources 
% (roughly, but not exactly, heap fragments). 
% Gordon~\cite{gordon2019modal} described a use of hybrid logic in the verification of actor programs, 
% where nominals named the local state of individual actors (with such assertions stabilized with a 
% rely/guarantee approach). 
Beyond these, there is limited work on the interaction of hybrid logic with substructural logics, in restricted forms
that do not affect expressivity~\cite{despeyroux2014hybrid,chaudhuri2019hybrid}.\looseness=-1
% Primarily there is a line of work on hybrid linear logic (\HyLL)~\cite{despeyroux2014hybrid}, 
% originally used as a way to more conveniently express aspects of transition systems in linear logic. 
% However, \HyLL's proof rules offer no non-trivial interactions with multiplicative connectives 
% (every \HyLL proof can in fact be embedded into regular linear logic~\cite{chaudhuri2019hybrid}, 
% unlike Brotherston and Villard's \HyBBI, which demonstrably increases expressive power over its base \BBI.
